---
title: "Visualizing DARTS CTD Data -- JMB Live Coding"
author: "Julia Brown"
date: "2026-09-17"
output: 
  pdf_document: default
  md_document:
    variant: gfm
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages that we will need to use for plotting data using R. The package is called "tidyverse" which is a huge set of tools that are great for data analysis and visualization. 


```{r}
library(tidyverse)
```
Now let's load our data frame into R. (the csv we just made)

```{r}
data <- read.csv("240905_darts_clean.csv")
```

What does our data look like? To see the structure of your data:

```{r}
str(data)
```

Plot something! temperature and depth:

```{r}
ggplot(data, aes(x=temp_c, y=depth_m)) + geom_point()
```

plotting water column temperature profile for station 1:

```{r}
stn1 <- filter(data, station == "S1")
```

== symbol tests for a value to be equivalent to what you provide
!= choose values not equal to what you provide 
> "greater than" can apply this test when the column values are integers or numeric values
< "less than" 

```{r}
ggplot(stn1, aes(x = temp_c, y = depth_m)) + 
  geom_point() + 
  scale_y_reverse() + 
  xlab('Temperature (C)') + 
  ylab('Depth (M)') + 
  ggtitle('Station 1 Temperature Profile')
```

```{r}
ggsave('stn1_temp_profile.png', width=4, heigh=6)
```

plotting profiles from multiple stations at the same time, in the same figure.

```{r}
ggplot(data, aes(x = temp_c, y = depth_m)) + 
  geom_point() + 
  facet_wrap(~station, ncol = 4) + 
  scale_y_reverse() + 
  ylim(100, 0) + 
  xlim(7.5, 18)
```

Plotting all profiles on the same plot:

```{r}
ggplot(data, aes(x = temp_c, y = depth_m, color = station)) + 
  geom_point(size = 1) + 
  scale_y_reverse() + 
  labs(color = 'Station')
```

JMB Scratch after class to play with plotting T-S diagrams for watermasses worksheet:

```{r}
ggplot(data, aes(y = temp_c, x = sal_psu, color = station)) + 
  geom_point(size = 1) + 
  scale_y_reverse() + 
  labs(color = 'Station')
```










