---
title: "1 Visualizing DaRTS CTD data from a single cruise"
author: "Julia Brown + Catherine Mitchell"
date: "2026-09-10"
output: 
  md_document:
    variant: gfm
  pdf_document: default
  html_document: default
  html_notebook: default
  
knit: (function(input, ...) {
    rmarkdown::render(input, output_format = "all", envir = globalenv())
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this lesson, we are going to explore CTD data from a single DaRTS cruise.

We will learn how to:

2.  Import a spreadsheet into R as a data frame.
3.  Plot variables by depth in R using ggplot2.

#### Scenario

David Fields has sent you an email, where he has attached an [excel file](https://docs.google.com/spreadsheets/d/1iZ4HbfRHGD-cPz6SCrBwKZYPxMUwdORl/edit?usp=sharing&ouid=100534300918978737444&rtpof=true&sd=true) from your most recent DaRTS cruise. His email is a little cryptic but says: 

```
Using R...
   1) plot TEMP, SALINITY, DENSITY, CHL, O2(%SATURATION) AND PH  in R with the Depth (m) on the vertical axis in reverse order.
```
    
Being a resourceful student, you open up the excel file to look at the data yourself, and find that it's not exactly ready to load into R.  *Why is this?* 

Some things you notice:

* **Embedded graphs**: It's nice that David added these pre-visualizations, but we won't be able to load these into R.

* **Multiple tabs**: Each station is represented by it's own tab in the spreadsheet. This will be a problem when we want to compare stations. It would be better to have all stations in the same table, with a column designating the station. 

* **Column headers**: The column headers have too many extra spaces and characters, which could cause problems when referencing them with R.

* **Dates**: There's no indication of when these data were collected outside of the name of the initial file. What if you wanted to merge data from multiple cruises?  A date column is important!

#### Cleaning Up the Data

Let's create a spreadsheet we will actually be able to load into R. We will start by creating a new tab in the Excel Spreadsheet. We will call this tab "combined". Then, we will add headers to the first row of this new table. The first two headers will be **date** and **station**, and then we will add headers for the measurement columns on the station tabs, but we will rename them slightly to be more coding/R friendly.

Here's how we will rename the measurement columns: 

  **Depth (M)** to "depth_m"   
  **Temp (deg C)** to "temp_c"  
  **Salinity (PSU)** to "sal_psu"  
  **Density (kg/m^3)** to "density_kg_per_m3"  
  **PAR/Irradiance** to "par"  
  **Fluorescence (mg/m^3)** to "fluor_mg_per_m3"  
  **Turbidity (NTU)** to "turbidity_ntu"  
  **Oxygen (umol/kg)** to "O2_umol_per_kg"  
  **Oxygen (% sat)** to "O2_pct_sat"  
  **Ph** to "ph"  
  
Next, we will copy and paste the data from each spreadsheet into this new table.  

Be sure to skip the header row when you copy and paste, and make sure that the first column you paste is below the "depth_m" column. 

With each tab, we will also need to make sure that the station number is transferred to the new spreadsheet in the "station" column, so when you copy and paste all the data from the "S1" tab, population the corresponding rows in the "station" column with "S1". 

The date will be 240905 throughout this whole spreadsheet, as all this data comes from the cruise that set out on that particular day.
  
Once you have all the data organized into one spreadsheet, you can export it as a csv to a known location on your computer.  A good place might be in a "darts_data" folder on your desktop. Let's name the file "240905_darts_clean.csv"

## Loading your data into R

Now that we have a clean spreadsheet, we can load it into R. Before that we'll want to load the R libraries we'll be using today. We're going to load a large package that has a lot of other packages within it, called the "tidyverse":

```{r}
library(tidyverse)
```
This loads all the functions we will need to load and plot our data.

Let's try to read in the file we created. The function to do this is `read.csv`, and we need to provide the name and location of our data file, including the path to the file if it is in a different location than this markdown file.

**Example 1:** not including the file path

`data <- read.csv("240905_darts_clean.csv", header = TRUE)`

**Example 2:** including the path

In the below `"~/Desktop"` indicates that you're starting in your home directory and then moving to your Desktop.
```{r}
data <- read.csv("./240905_darts_clean.csv", header = TRUE)
```

You can see this object show up on the left-hand 'Environment' tab in R studio.  You can learn more about this object using a couple of different commands. For example, to learn about what the structure of the object is you can type use the function 'str'

```{r}
str(data)
```

Other functions you can use to learn about the data object are "colnames" and "nrow".  Feel free to give those a try to see what they do.

### Visualizing Data: Part 1

**Challenge 1:** Plot a temperature profile from Station 1.

Let's think through the steps we're going to need to do before we start:

1. Select only the data from Station 1
2. Plot temperature on the x-axis, depth on the y-axis

#### Wrangling Data 
##### *aka getting your data ready for plotting*

We need to pull out all rows of data that correspond to Station 1.  We will do this using a process called *filtering*. We will use the function `filter` from the package `dplyr` (part of the tinyverse). It looks like this:

```{r}
stn1 <- dplyr::filter(data, station == "S1")
```

**What did we just do?** The double equals sign is a *conditional*, it's telling R to go through the data frame and identify rows where the condition of equality is met for the specified variables (in this case, where the station column equaled 1).

It can be good practice to check what you did worked - so let's check what this looks like (using code below, or could check the `stn1` variable in the Environment):
```{r}
head(stn1)
```

Now we are ready to plot our data.

**But what about filtering out the temperature and depth data?? Don't we need to do that first?**  No!

When plotting, we can tell R which columns of our data frame we'd like it to plot, and it will plot all the rows in the data frame for those columns. Our data frame contains rows for all four stations. So if we want to plot only one station, we need to filter out the station we want first. Then we tell R to plot the columns we want from that filtered data frame.

## Plotting Data

Time to plot! We're going to use a package called `ggplot2`... we will go into greater depth later about how exactly ggplot is working, for now, here's the code to plot a temperature profile:

```{r}
ggplot(stn1, aes(x = temp_c, y = depth_m)) + geom_point()
```

How does this look? Is there anything wrong with this plot?

In oceanography, we like to orient ocean data from the surface on the top of the y axis and depth at the bottom of the y axis, so we need to flip our plot (aka reverse our y axis) - note, part of programming is learning the language to use:

```{r}
ggplot(stn1, aes(x = temp_c, y = depth_m)) + 
  geom_point() + 
  scale_y_reverse()
```

R has automatically named our axes using the data frame column names. This is super handy for quick plotting, but often we want to relabel them into something more informative:

```{r}
ggplot(stn1, aes(x = temp_c, y = depth_m)) + 
  geom_point() + 
  scale_y_reverse() + 
  xlab('Temperature (C)') + 
  ylab('Depth (m)') + 
  ggtitle('Station 1 Temperature Profile')
```

## Saving figures

If you want to save a figure, use `ggsave`:
```{r}
ggsave('st1_temp_profile.png', width = 4, height = 6)
```

Although the nice thing about using notebooks and markdown is that you don't need to save your figures out - they are rendered within your markdown or notebook file.

## Visualizing Data: Part 2

**Challenge 2:** Plot a temperature profile from all four stations.

Let's think through the steps we're going to need to do before we start:

In this case, we have a few options:
* Do we want four individual plots, similar to what we just created?
* Do we want all the data shown on one plot?

It probably depends the story you're trying to tell, or what the data looks like, or some other factor.

Let's do both!

### Individual plots

Here, we do NOT need to redo what we did above for station 1 four times. We can use a technique called `faceting`:

```{r}
ggplot(data, aes(x = temp_c, y = depth_m)) +
  geom_point() + 
  facet_wrap(~station, ncol = 4) +
  scale_y_reverse()
```
Note that you can specify the axis limits:

```{r}
ggplot(data, aes(x = temp_c, y = depth_m)) +
  geom_point() + 
  facet_wrap(~station, ncol = 4) +
  ylim(100, 0) + xlim(7.5, 18)
```

### Single plot for all stations

Here, we add an additional argument to the call to `aes` to plot each station's data a different:

```{r}
ggplot(data, aes(x = temp_c, y = depth_m, color = factor(station))) +
  geom_point(size = 1) + 
  scale_y_reverse() + 
  labs(color = 'Station')
```

Note - let's breakdown that `factor(station)` command:

*`factor(station)` is doing something to our station column... it's turning those numbers into factors. What is a factor? Factors are categories i.e., they are discrete buckets, rather than continuous data. So, because our stations can only be 1, 2, 3 or 4, they are factors, rather than regular numbers, because if they were regular numbers, it would imply we could have a station 1.1, 2.5, etc. 
* The conversion to factors is important for how R considers the data. If R thinks the data is a continuous number, it will color the dots differently - see example below.

If you don't do the conversion to factors:
```{r}
ggplot(data, aes(x = temp_c, y = depth_m, color = station)) +
  geom_point(size = 1) + 
  scale_y_reverse() + 
  labs(color = 'Station')
```

# Practice Questions    

1. Plot salinity, density and O2 profiles for each station using faceting
2. For each of the following variables, plot profiles for each station on one plot: PAR, fluorescence, turbidity and O2 percent saturation (i.e., one plot for PAR, with all four stations on the one plot, etc).
   
In both cases, be sure to label the axes appropriately (with the correct units). 

Bonus questions:

a. Can you change the color, shape and/or size of the dots in the facet plot?
b. Can you change the color, shape and/or size of the dots in the single plot?


