---
title: "Data-prep-for-TS-worksheet"
output: 
  pdf_document: default
  github_document: default
  html_document: default
  html_notebook: default
always_allow_html: true
---

Using Claude to help find some AMT profiles for TS worksheet. We'll identify appropriate CTD casts, and make a csv file containing our selected profiles.

```{r}
library(tidyverse)
```


```{r}
# Point this at wherever you unzipped the NCEI download
parent_dir <- "C:/Users/cmitchell/Documents/SeaChangeSemester/2026/RLabs/data/"
data_dir <- paste0(parent_dir,"0044519")
```

What's in the CSV? We know it's come from NCEI and the format these files generally are:

```{r}
# List everything so you can see how it's organized (per-cast files vs one big file)
list.files(data_dir, recursive = TRUE) %>% head(20)

# Peek at one file's raw structure before trying to parse it
read_lines(list.files(data_dir, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)[1],
           n_max = 15)
```
OK - from this output, we can write functions to read in the data:

```{r}
parse_dms <- function(x) {
  # e.g. "36 36 50N" or "017 30 18W" -> decimal degrees
  hemi <- str_sub(x, -1)
  nums <- as.numeric(str_extract_all(x, "\\d+\\.?\\d*")[[1]])
  dd <- nums[1] + nums[2] / 60 + ifelse(length(nums) >= 3, nums[3], 0) / 3600
  if (hemi %in% c("S", "W")) dd <- -dd
  dd
}

read_amt_cast <- function(path) {
  raw <- read_lines(path)

  meta_names <- str_split(raw[1], ",")[[1]] %>% str_trim()
  meta_vals  <- str_split(raw[2], ",")[[1]] %>% str_trim()
  meta <- set_names(as.list(meta_vals), meta_names)

  param_codes <- str_split(raw[3], ",")[[1]] %>% str_trim()
  param_codes <- param_codes[param_codes != ""]

  col_names <- as.vector(rbind(param_codes, paste0("flag_", param_codes)))

  df <- read_csv(path, skip = 4, col_names = FALSE, show_col_types = FALSE,
                  na = c("", "NA"))
  df <- df[, seq_along(col_names)]
  names(df) <- col_names

  # Pull a column by name if it exists, otherwise return a column of NA
  get_col <- function(df, name) {
    if (name %in% names(df)) df[[name]] else rep(NA_real_, nrow(df))
  }

  tibble(
    depth = get_col(df, "PRESPR01"),
    temperature = get_col(df, "TEMPST01"),
    salinity = get_col(df, "PSALST01"),
    chlorophyll = get_col(df, "CPHLPR01"),
    cruise = meta$Cruise,
    cast_id = meta$Orig._Id,
    latitude = parse_dms(meta$Latitude),
    longitude = parse_dms(meta$Longitude),
    start_date = meta$Start_Date
  )
}
```

Now we can read the data, and find where there's chl data too:

```{r}
cast_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)

all_casts <- map_dfr(cast_files, read_amt_cast)

# keep casts with real (non-zero, non-NA) chlorophyll data
casts_with_chl <- all_casts %>%
  group_by(cast_id) %>%
  filter(any(!is.na(chlorophyll) & chlorophyll > 0)) %>%
  ungroup()
```

Let's check which cruises have chl profiles:
```{r}
cruise_summary <- casts_with_chl %>%
  distinct(cruise, cast_id, latitude) %>%
  group_by(cruise) %>%
  summarise(
    n_casts = n(),
    lat_range = max(latitude) - min(latitude),
    .groups = "drop"
  ) %>%
  arrange(desc(n_casts))

cruise_summary
```
Let's select our cruise:
```{r}
chosen_cruise <- "AMT12"  # swap in whichever cruise looks best from Step 1

casts_deep_enough <- casts_with_chl %>%
  group_by(cast_id) %>%
  filter(max(depth, na.rm = TRUE) > 500) %>%
  ungroup()

cruise_casts <- casts_deep_enough %>%
  filter(cruise == chosen_cruise) %>%
  distinct(cast_id, latitude) %>%
  arrange(latitude)

cruise_casts  # inspect the actual latitude spread available
```

And now we can select our profiles:
```{r}
# pick the cast nearest each target latitude
target_lats <- c(45, 25, 0, -25, -50)

chosen_ids <- map_dfr(target_lats, function(t) {
  cruise_casts %>%
    distinct(cast_id, latitude) %>%
    mutate(dist = abs(latitude - t)) %>%
    slice_min(dist, n = 1)
})

five_profiles <- casts_with_chl %>% filter(cast_id %in% chosen_ids$cast_id)
```


And finally, plotting:

```{r}
ggplot(five_profiles, aes(salinity, temperature, color = factor(round(latitude)))) +
  geom_path(aes(group = cast_id), linewidth = 0.6) +
  geom_point(size = 1) +
  labs(x = "Salinity (PSU)", y = "Temperature (°C)", color = "Latitude",
       title = "T-S diagram: 5 AMT profiles across the Atlantic") +
  theme_minimal()
```

plotting by chl:

```{r}
ggplot(five_profiles, aes(salinity, temperature)) +
  geom_path(aes(group = cast_id), color = "grey60", linewidth = 0.4) +
  geom_point(aes(color = chlorophyll), size = 1.8) +
  scale_color_viridis_c(name = "Chl-a\n(mg/m³)", trans = "log10") +
  labs(
    x = "Salinity (PSU)",
    y = "Temperature (°C)",
    title = paste("T-S diagram:", chosen_cruise, "— 5 profiles across the transect"),
    subtitle = "Points colored by chlorophyll-a concentration"
  ) +
  theme_minimal()
```

```{r}
ggplot(five_profiles, aes(salinity, temperature)) +
  geom_path(aes(group = cast_id), color = "grey60", linewidth = 0.4) +
  geom_point(aes(color = depth), size = 1.8) +
  scale_color_viridis_c(name = "Depth (m)", trans = "reverse") +
  labs(
    x = "Salinity (PSU)",
    y = "Temperature (°C)",
    title = paste("T-S diagram:", chosen_cruise, "— 5 profiles across the transect"),
    subtitle = "Points colored by depth"
  ) +
  theme_minimal()
```

Saving the file:
```{r}
write_csv(five_profiles, paste0(parent_dir,"amt_five_profiles.csv"))
```
